Smart Representations: {R}ationality and Evolution in a Richer Environment

### 2015-12-18 paper/paper.tex
sec com

A \emph{choice mechanism} is a map from choice situations to action choices that, unlike the
classic rules visited above, contains an explicit level of subjective representation of a
decision situation. Let us think of a choice mechanism as a pair of functions
($\gamma,\kappa$), where $\gamma$ is a \emph{subjective representation scheme} and $\kappa$ is
an \emph{action selection function} (see Figure~\ref{fig:ChoiceMechanism}).

The subjective representation scheme $\gamma_i$ of agent $i$ is a function $\gamma_i$ that
takes a strategic game $G = \tuple{ N, (A_i , \pi_i)_{i \in N} }$ and maps it onto
$\gamma_i(G) = (\theta, e_i)$ where
$\theta \mycolon \bigtimes_{i \in N} A_i \rightarrow \mathbb{R}$ is a (subjective) payoff
function for agent $i$ suitable for $G$.

$\mathcal{G} \rightarrow \Theta \times E$, where $\mathcal{G}$ is a class of

Formally, given a class of strategic games $\mathcal{G}$, the subjective representation scheme
for agent $i$ is a function $\gamma_i: \mathcal{G} \rightarrow \Theta \times E$ that gives a
subjective preference $\theta \in \Theta$ and an epistemic state $e \in E$ for any game
$G \in \mathcal{G}$. A subjective preference $\theta$ for $G$ and $i$ is just a utility
function for $G$. This can but need not be different from the actual payoff function of $G$. An
epistemic state $e$ for $G$ and $i$ is some specification of $i$'s uncertainty about the
co-players' behavior.

The action selection function $\kappa$ maps such a subjective representation onto a concrete
action choice. An action selection function for agent $i$ is a function
$\kappa_i: \Theta \times E \rightarrow \pow{A_i} $, that associates any subjective
representation $\gamma_i(G) = (\theta,e)$ of $G$ with a set of prescribed actions for $i$ in
$G$ under $i$'s subjective representation of $G$. We drop the index $i$ for readability in the
following.

To keep matters simple the focus is on \emph{uniform} and \emph{general} ways of forming
preferences and beliefs consistently across decision situations. We consider two
\emph{preference types} and two \emph{belief types}: subjective preferences either coincide
with the actual, true payoffs or are given by the regret transformation of the actual game;
beliefs are either precise probabilistic beliefs modelled by a single probability distribution,
or they are imprecise probabilistic beliefs modelled by a set of probability distributions. The
action selection function $\kappa$ is fixed to \textit{maxmin expected utility}
\citep{gilsch89}, for reasons explained in the following.

% It turns out that whether $\tau^{reg}$ can have a fitness advantage over $\tau^\pi$ depends on
% the general \emph{kind} of (probabilistic) beliefs that DM's form about the opponent's
% behavior. In other words, we even abstract away from the content of the DM's beliefs about his
% opponents' behavior that he could have acquiresd by learning or observation, but first turn on
% the more general question of how uncertainty should be represented in order to interact with
% subjective preferences in such a way that the resulting behavior will be fitness enhancing for
% the DM in the long run. We consider therefore two \emph{belief types} (to be defined below):
% imprecise belief types $\Delta(A_{-i})$, who have a set of probabilistic beliefs about the
% opponents' behavior and precise belief types $\overline{\mu}$ who have a single probabilistic
% belief.

% The following subsection describes what these belief types are and how a single action
% selection function (maxmopin expected utility) gives rise to an intersting parallel with
% classic choice rules.

@@

\subsection{Choice mechanisms \& classic choice rules}

By fixing the action selection component to maxmin expected utility, and allowing two belief and
two preference types, we regain all the choice rules considered in sectionNN from a more general and unified standpoint, as it is summed up in
Table~\ref{tab:CMs}. \\
We are now in the position of formally deriving behavior from a pair $(\theta, e) \in \Theta \times E$.

\begin{definition} \label{def:utility-belief}

Given a game $G$, an action $a_i \in A_i$, and a belief $\mu \in \lbrace \mu' \in \mathbb{R}^{A_{-i}}: \sum_{a_{-i} \in A_{-i}} \mu'(a_{-i}) = 1 \rbrace$, we define $\tau(G)(a_i, \mu):= \sum_{a_{-i} \in A_{-i}} \tau(G)(a_i, a_{-i}) \cdot \mu(a_{-i})$.

\begin{definition} \label{def:maxminSEU}

Given a game $G$, a preference type $\tau$ and a belief type $e$, we define $\kappa(\tau(G), e):= \lbrace a_i \in A_i : a_i \in \text{max}_{a'_i \in A_i} \text{ min}_{\mu \in e} \text{ } \tau(G)(a'_i, \mu) \rbrace$.\footnote{As already specified in the previous example, we assume that, whenever the set $\kappa(\tau(G), e)$ is not a singleton, the player $(\tau, e)$ will mix with equal probability among the actions in $\kappa(\tau(G), e)$.}

It is then worth stressing that $\kappa(\tau^{\pi}, \Delta)$ turns out to be the actual payoff
maximinimizer, $\kappa(\tau^{reg}, \Delta)$ corresponds to the regret minimizer,
$\kappa(\tau^{\pi}, \overline{\mu})$ is the expected actual payoff maximizer (with flat beliefs), and
$\kappa(\tau^{reg}, \overline{\mu})$ is the expected regret minimizer (with flat beliefs).
Moreover, different subjective preferences and belief types can give rise to identical action predictions,
either in a restricted class of games, or in general. This is also the case here.

\begin{fact} \label{fact:maxEU-minReg} For any strategic game $G$ and any arbitrary
  probabilistic belief $\mu \in \Delta$ about the opponents' behavior, the acts
  selected by maximization of expected utility for $G$ under $\mu$ are exactly the acts
  selected by the maximization of expected utility for the negative regret transformation of $G$
  under $\mu$ \citep[e.g.,][]{HalpernPass2012:Iterated-Regret}.
\end{fact}

\noindent In other words, given the same probabilistic belief, maximization of expected
(objective) payoff and minimization of expected regret are behaviorally
equivalent. This holds for arbitrary games, as well as for solitary decision situations. On the
other hand, the following does not hold in general:

\begin{fact} \label{fact:equivalence2x2}
In the class of $2 \times 2$ symmetric games, players $(\tau^{reg}, \Delta)$, $(\tau^{\pi}, \overline{\mu})$, and $(\tau^{reg},\overline{\mu})$ are behaviorally equivalent.
\end{fact}

\begin{example}
  Consider the game in Figure \ref{coordgame1} again. When a player characterized
  by\footnote{To ease notation in the following we will refer to the pair
    $(\text{preference type } \tau,\text{ belief type } e)$ as the \textit{player type}.} preference
  type $ \tau^{reg} $ and belief type $ \Delta $ faces the game in Figure
  \ref{coordgame1}, her preference type gives rise to the representation $\tau^{reg}(G^C)$ in
  Figure \ref{coordgame1reg}. Then, maximinimizing expected utility over the probability
  simplex $ \Delta $ given the game in Figure \ref{coordgame1reg} corresponds to
  playing the standard maxmin strategy [REF] of that game.\footnote{Notice that this is the
    case for any game: given a game $G$, maximinimizing expected utility over the probability
    simplex $ \Delta $ amounts to playing the standard maxmin strategy of $G$.} This,
  in turn, corresponds to the regret minimizing action of the original game $G^C$ of Figure
  \ref{coordgame1}, namely action $II$, as we have seen in Section \ref{sec:choice-rules}. A
  player $(\tau^{reg},\overline{\mu})$ instead has subjective representation $\tau^{reg}(G^C)$
  of Figure \ref{coordgame1reg} and a flat probability measure
  $\overline{\mu}=(\frac{1}{2}I,\frac{1}{2}II)$ over the co-player's actions. Consequently,
  such a player would play action $II$, i.e., the action that maximinimizes expected utility
  given the flat probability measure $\overline{\mu}$. Similarly, a player
  $(\tau^{\pi},\Delta)$ will have the subjective representation coinciding with the
  objective fitness game, i.e., $ \tau^{\pi}(G^C)= G^C $. Then, both actions $I$ and $II$
  maximinimize expected utility over the probability simplex $ \Delta $. Hence, a
  player $(\tau^{\pi},\Delta)$ will be indifferent between $I$ and $II$, and will
  randomize with equal probability $(\frac{1}{2}I,\frac{1}{2}II)$. Finally, a player
  $(\tau^{\pi},\overline{\mu})$ maximinimizes the objective fitness game $G^C$ of Figure
  \ref{coordgame1} with flat probability $\overline{\mu}$ over the co-player's actions, i.e.,
  maximizes objective utility given $\overline{\mu}$. In game $G^C$ this yields action $II$.

\end{example}

%\todo[inline]{should we explain here, or very soon, what the resulting types really are: regret minimizers, maxmin players \dots}

Many other alternatives different from maxmin could produce behavior: choice rules like minmax, maxmax, and minmin would similarly generate behavior in case of a lack of precise probabilistic belief. As it should be clear by now, here we postulate that players use maxmin for two main reasons. First, among other possible choice rules maxmin is not only intuitively appealing but also corroborated from an empirical point of view. Secondly, our main interest is the evolution of different patterns of subjective conceptualizations of games, and we want to show that the veridical representation is not necessarily the fittest and may not be evolutionarily stable. Assuming that all players use a fixed rule (maxmin expected utility) independent of their preference and belief type allows us to factor the choice rule out of the selection process and to focus on subjective representations only.

\begin{fact} \label{fact:maxEU-minReg}

For any probability measure $\mu$, maximization of expected utility and minimization of expected regret are behaviorally equivalent.

\end{fact}

\noindent As we have seen in the previous example, this is not the case in the presence of (radical) uncertainty, in the sense of representation of belief state in terms of a set of probability, as in $ \Delta $.
\fi

>>>>

A \emph{choice mechanism} is a map from choice situations to action choices that, unlike the
classic choice rules visited above, contains an explicit level of subjective representation of
a decision situation. Let us think of a choice mechanism as a pair of functions
($\gamma,\kappa$), where $\gamma$ is a \emph{subjective representation scheme} and $\kappa$ is
an \emph{action selection function} (see Figure~\ref{fig:ChoiceMechanism}).

The subjective representation scheme $\gamma_i$ of agent $i$ is a function $\gamma_i$ that
takes a strategic game $G = \tuple{ N, (A_i , \pi_i)_{i \in N} }$ and maps it onto
$\gamma_i(G) = (\theta, e_i)$ where
$\theta \mycolon \bigtimes_{i \in N} A_i \rightarrow \mathbb{R}$ is payoff function suitable
for $G$, and $e_i$ is some representation of $i$'s uncertainty about the co-player's
behavior. We consider two \emph{preference types}: the \emph{objective type} maps each game
onto its own, true payoff function; The \emph{regret type} maps each game onto its regret
transformation, as introduced above. We also consider two \emph{belief types}: the
\emph{precise type} represents uncertainty in a single probability distribution
$\mu_i \in \Delta(\Acts_{-i})$ about the co-players' behavior; the \emph{imprecise type}
represents uncertainty as a set of probability distributions $X \subseteq \Delta(\Acts_{-i})$.
It is convenient to conceptualize precise beliefs as a special case of imprecise beliefs by
looking at $\mu_i$ as a stand-in for the singleton set $\set{\mu_i}$. More details about
imprecise probabilities and our initial assumption of ignorance are given in
Section~\ref{sec:impr-prob-beli} below.

The only action choice function $\kappa$ that we consider here is \emph{maximinimization of
  expected (subjective) utilities} (see Section~\ref{sec:impr-prob-beli} for
motivation).

\begin{definition}[Maximinimization of expected (subjective) utility]
  \label{def:maxminSEU}
  Fix a strategic game $G$. Let agent $i$'s subjective representation scheme
  $\gamma_i(G) = (\theta, e_i)$ be such that $\theta$ is $i$'s subjective payoff function for
  $G$ and $e_i$ is a (possibly singleton) set of probability distributions in
  $\Delta(\Acts_{-i})$. Agent $i$'s expected (subjetive) utility for act $\myact{i}$ and
  $\mu \in e$ is
  $\EU(\myact{i}, \mu) = \sum_{\myact{-i} \in \Acts_{-i}} \mu(\myact{-i}) \theta(\myact{i},
  \myact{-i})$.
  Act $\myact{i}^*$ is selected by maximinimization of expected (subjective) utility iff:
  $\myact{i}^* \in \arg \max_{\myact{i} \in \Acts_{i}} \min_{\mu \in e} \EU(\myact{i}, \mu)$.
\end{definition}

Maximinimization under a precise (singleton set) probabilistic belief is just standard
maximization of expected utilities. Moreover, given the same precise (singleton set)
probabilistic belief, the choices of objective and regret preference types are exactly the
same.

\begin{fact} \label{fact:maxEU-minReg} For any strategic game $G$ and any arbitrary precise
  (singleton set) probabilistic belief about the opponents' behavior, the acts selected by
  maximization of expected utility for $G$ are exactly the acts selected by the maximization of
  expected utility for the negative regret transformation of $G$
  \citep[e.g.,][]{HalpernPass2012:Iterated-Regret}. This also holds for solitary decision
  problems.
\end{fact}

This means that as far as behavior is concerned, regret types cannot be distinguished from
objective types, as long as both hold precise beliefs. This will be central in understanding
later results: under precise beliefs evolution will not favor either preference type over the
other. On the other hand, if paired with imprecise beliefs, different behavioral predictions
arise. However, in the special case of $2 \times 2$ symmetric games, another noteworthy general
result holds.

\begin{fact} \label{fact:equivalence2x2} In the class of $2 \times 2$ symmetric games, acts
  selected by Laplace's rule are exactly the acts selected by regret minimization. 
\end{fact}

In the following, we assume that agents are maximally ignorant, so that precise types have a
(singleton set containing a) flat probability distribution
$\overline{\mu}_i \in \Delta(\Acts_{-i})$ and imprecise types consider all probabilistic
beliefs about the co-players' behavior $\overline{X} = \Delta(\Acts_{-i})$. In that case, the
considered choice mechanisms are behaviorally equivalent to classic choice rules, as summarized
in Table~\ref{tab:CMs}. This means that most of the following results which we would like to
interpret as results about the evolutionary fitness of subjective representation schemes have
value also for anyone interested only in a comparison of the ecological value of classic choice
rules. This also means that, given Fact~\ref{fact:equivalence2x2}, all choice mechanisms are
behaviorally equivalent on $2 \times 2$ symmetric games, except for objective preference types
with imprecise beliefs.

@@
